Modelling the emergence of whisker barrels

### 2020-06-03 paper/briefpaper/brief.tex
Results of running model 10 times to get variation in outputs based on noisy initial conditions

The only differences between thalamic projections are their strengths of
interaction with the guidance fields, $\gamma$, hence any reliable differences
between emergent cortical fields must be due to differences in these values
only. We speculate that the contribution of a given ephrin field, to the
velocity at which a projection migrates across the cortical subplate, is
determined by the concentration of a similar molecule at its thalamic origin,
i.e., the putative barreloid center. \cmnt{Thus, we suggest that there exist a
  pair of signalling molecules whose expression varies linearly (with
  orthogonal gradients) across the thalamus, analgous to the ephrins, whose
  existence is well known in the cortex. Thalamocortical axons growing from a
  region in the thalamus with high expression of one of these molecules will
  interact with a corresponding cortical signalling molecule more strongly
  than an axon emanating from a region with low expression of the thalamic
  signal.}  \cmnt{We defined two orthogonal, linear gradients oriented in the
  anterior-posterior and medial-lateral directions and used the coordinates of
  $41$ barreloid centers recreated from \cite{haidarliu_size_2001},
  corresponding to the macrovibrissae, to populate 41 pairs of $\gamma$
  values.}  (Fig.\,\ref{fig:main}B). \cmnt{In other words, we simply scaled
  the barreloid coordinates in an anterior-posterior/medial-lateral coordinate
  system to create the $\gamma$ values. Note that positive \emph{and} negative
  $\gamma$ values were generated; the origin of the coordinate system was
  located at the center of the thalamic barreloid field. This prevented an
  overall drift of the cortical pattern away from the centroid of the barrel
  field and may explain why coordinate-giving transcription factors and the
  ephrins are often found as pairs acting in opposing directions.}
\cmnt{We concentrated on macrovibrissae (excluding microvibrissae) because it
  is easy to identify their corresponding structures both in the thalamus and
  in the cortex. We also excluded the 5 rhinal whiskers because their cortical
  representation is in a separate island, distinct, and somewhat removed from
  the representation of the macrovibirissae.}

A cortical boundary enclosing 41 \cmnt{macrovibrissae} barrels was traced from
a cytochrome oxidase stain from \cite{zheng_signal_2001} \cmnt{(using the
  authors' original data)}, and Eqs.~\ref{eq:dc}--\ref{eq:comp} were solved
for $N=41$ projections on the resulting domain, using $M=2$ linear signalling
gradients aligned with the anterior-posterior and medial-lateral axes. From
random initial conditions for $a(\mb{x},0)$ and $c(\mb{x},0)$, a clear
Voronoi-like tesselation emerged (Fig.\,\ref{fig:main}C; see Movie S1). A
reduction in the Honda delta metric (see \citealp{senft_mouse_1991}) confirmed
that a `good' Voronoi pattern emerged within \cmnt{$\approx$~20000} iterations
(Fig.\,\ref{fig:main}D).  A measure of the difference in \cmnt{barrel area}
between real and simulated barrels revealed a strong correspondence (see
Fig.\,\ref{fig:main}D, squares). \cmnt{In a chosen region, the ratio of
  \emph{the connection density of the TC type $i$ which makes the densest
    connections} to \emph{the sum of all connection densities} gives a measure
  of the localization of the connection density. We computed this measure
  across the entire barrel domain and named it $\omega$ (see
  Fig.\,\ref{fig:main}D, grey hexagons). $\omega$ increases during the first
  20000 simulation steps and converges on a stable
  value. Fig.\,\ref{fig:main}E shows the localization measure plotted as a
  luminance map, giving a more graded impression of the structure which
  develops in the system.}

% Full page width
\begin{figure}
  \begin{fullwidth}
    \includegraphics[width=\linewidth]{./Fig2.png}
    \caption{\cmnt{\textbf{A} Exploring the parameter space of the model given
        in Eqs.\,1--3. Graphs show the three pattern metrics for $t=30000$
        steps as colourmaps: Honda $\delta$ (top row) shows conformance to a
        Voronoi pattern; $\eta$ (middle row) gives the simulated/experimental
        barrel area difference and $\omega$ (bottom row) is a measure of
        localization (separation) of the connection densities for the 41
        thalamocortical axon types. For each row, lighter (orange/yellow)
        colours imply a better model. The parameter space explored is three
        dimensional with the parameter ratio $\alpha/\beta$ varied along the
        columns of graphs, the competition parameter $\epsilon$ varied on the
        $y$-axes and the diffusion constant, $D$ varied along the $x$-axes. An
        asterisk (*) marks the parameter set displayed in
        Fig.\,\ref{fig:main}. Boxes i), ii) and iii) mark parameter sets for which
        corresponding patterns are shown in \textbf{B} (for $t=30000$ steps).}}
    \label{fig:paramsweep}
  \end{fullwidth}
\end{figure}

@@

\cmnt{The maps indicate that the pattern is not strongly influenced by the
  choice of $\alpha/\beta$, which was varied by 5 orders of magnitude. The
  most Voronoi-like patterns were achieved for parameter sets for which $D$
  and $\epsilon$ were high. The barrel area metric, $\eta$ indicates a strong
  preference for $D\approx0.5$, although this does not give the maximally
  localized solution, which for $D$ should be small. Thus, there is a
  trade-off between good localization and good Voroni pattern
  conformation. The
  parameter set indicated by the asterisk in
  Fig.\,\ref{fig:paramsweep}A is presented in Fig.\,\ref{fig:main} because it
  provides the best barrel area correspondence and produces a good Voronoi
  pattern. In Fig.\,\ref{fig:paramsweep}B, we present some patterns
  representing alternative parameter choices. Fig.\,\ref{fig:paramsweep}B i)
  shows the simulation result with low diffusion constant, $D$. Reducing the
  diffusion constant is equivalent to increasing the size of the domain $S$
  and so this simulation can be thought to represent an animal with a large
  cortical area but with unchanged cortical complexity. In this simulation, a
  large proportion of the domain is occupied by the projections with
  intermediate interaction parameters; those with strong interaction
  parameters are bunched up towards the edge of the domain. A barrel pattern
  fails to form. Fig.\,\ref{fig:paramsweep}B ii) and iii) indicate the effect
  of varying $D$ in either direction from its optimum. ii) is a less
  exaggerated version of the pattern in i), whereas iii) shows an expansion of
  the outer barrels and a corresponding compression of the barrels within.}

@@

The present results suggest that the key requirements for the emergence of
realistic barrel patterning are i) at each cortical location thalamocortical
projections compete for a limited number of available synaptic connections
(Eqs.~\ref{eq:dc}--\ref{eq:da}), ii) at each location the branching rate of a
given projection is reduced by the density of other projections
(Eq.\,\ref{eq:comp}), and iii) the branch density of each projection is
conserved over time.

>>>>

The only differences between thalamic projections are their strengths of
interaction with the guidance fields, $\gamma$, hence any reliable differences
between emergent cortical fields must be due to differences in these values
only. We speculate that the contribution of a given ephrin field, to the
velocity at which a projection migrates across the cortical subplate, is
determined by the concentration of a similar molecule at its thalamic origin,
i.e., the putative barreloid center. \cmnt{Thus, we suggest that there exist a
  pair of signalling molecules whose expression varies linearly (with
  orthogonal gradients) across the thalamus, analgous to the ephrins, whose
  existence is well known in the cortex. Thalamocortical axons growing from a
  region in the thalamus with high expression of one of these molecules will
  interact with a corresponding cortical signalling molecule more strongly
  than an axon emanating from a region with low expression of the thalamic
  signal.}  \cmnt{We defined two orthogonal, linear gradients oriented in the
  anterior-posterior and medial-lateral directions and used the coordinates of
  41 barreloid centers recreated from \cite{haidarliu_size_2001},
  corresponding to the macrovibrissae, to populate 41 pairs of $\gamma$
  values.}  (Fig.\,\ref{fig:main}B). \cmnt{In other words, we simply scaled
  the barreloid coordinates in an anterior-posterior/medial-lateral coordinate
  system to create the $\gamma$ values in the range [-2,2]. Note that positive
  \emph{and} negative $\gamma$ values were generated; the origin of the
  coordinate system was located at the approximate center of the thalamic
  barreloid field. This prevented an overall drift of the cortical pattern
  away from the centroid of the barrel field and may explain why
  coordinate-giving transcription factors and the ephrins are often found as
  pairs acting in opposing directions.}  \cmnt{We concentrated on
  macrovibrissae (excluding microvibrissae) because it is easy to identify
  their corresponding structures both in the thalamus and in the cortex. We
  also excluded the 5 rhinal whiskers because their cortical representation is
  in a separate island, distinct, and somewhat removed from the representation
  of the macrovibirissae.}

A cortical boundary enclosing 41 \cmnt{macrovibrissae} barrels was traced from
a cytochrome oxidase stain from \cite{zheng_signal_2001} \cmnt{(using the
  authors' original data)}, and Eqs.~\ref{eq:dc}--\ref{eq:comp} were solved
for $N=41$ projections on the resulting domain, using $M=2$ linear signalling
gradients aligned with the anterior-posterior and medial-lateral axes. From
random initial conditions for $a(\mb{x},0)$, \cmnt{with each hexagonal element
  of $a_i$ set from a uniform distribution according to $0.2+{U}(0,0.2)$, and
  $c(\mb{x},0)=0$}, a clear Voronoi-like tesselation
emerged (Fig.\,\ref{fig:main}C; see Movie S1). A reduction in the Honda delta
metric (see \citealp{senft_mouse_1991}) confirmed that a `good' Voronoi
pattern emerged within \cmnt{$\approx$~20000} iterations
(Fig.\,\ref{fig:main}D).  A measure of the difference in \cmnt{barrel area}
between real and simulated barrels revealed a strong correspondence (see
Fig.\,\ref{fig:main}D, squares). \cmnt{In a chosen region, the ratio of
  \emph{the connection density of the TC type $i$ which makes the densest
    connections} to \emph{the sum of all connection densities} gives a measure
  of the localization of the connection density. We computed this measure
  across the entire barrel domain and named it $\omega$ (see
  Fig.\,\ref{fig:main}D, grey hexagons). $\omega$ increases during the first
  20000 simulation steps and converges on a stable
  value. Fig.\,\ref{fig:main}E shows the localization measure plotted as a
  luminance map, giving a more graded impression of the structure which
  develops in the system.}

% Full page width
\begin{figure}
  \begin{fullwidth}
    \includegraphics[width=\linewidth]{./Fig2.png}
    \caption{\cmnt{\textbf{A} Exploring the parameter space of the model given
        in Eqs.\,1--3. Graphs show the three pattern metrics for $t=30000$
        steps as colourmaps: Honda $\delta$ (top row) shows conformance to a
        Voronoi pattern; $\eta$ (middle row) gives the simulated/experimental
        barrel area difference and $\omega$ (bottom row) is a measure of
        localization (separation) of the connection densities for the 41
        thalamocortical axon types. For each row, lighter (orange/yellow)
        colours imply a better model. The parameter space explored is three
        dimensional with the parameter ratio $\alpha/\beta$ varied along the
        columns of graphs, the competition parameter $\epsilon$ varied on the
        $y$-axes and the diffusion constant, $D$ varied along the $x$-axes. An
        asterisk (*) marks the parameter set displayed in
        Fig.\,\ref{fig:main}. Boxes i), ii)~and iii)~mark parameter sets for which
        corresponding patterns are shown in \textbf{B} (for $t=30000$ steps).}}
    \label{fig:paramsweep}
  \end{fullwidth}
\end{figure}

@@

\cmnt{The maps indicate that the pattern is not strongly influenced by the
  choice of $\alpha/\beta$, which was varied by 5 orders of magnitude. The
  most Voronoi-like patterns were achieved for parameter sets for which $D$
  and $\epsilon$ were high. The barrel area metric, $\eta$ indicates a strong
  preference for $D\approx0.5$, although this does not give the maximally
  localized solution, which for $D$ should be small. Thus, there is a
  trade-off between good localization and good Voroni pattern
  conformation. The
  parameter set indicated by the asterisk in
  Fig.\,\ref{fig:paramsweep}A is presented in Fig.\,\ref{fig:main} because it
  provides the best barrel area correspondence and produces a good Voronoi
  pattern. In Fig.\,\ref{fig:paramsweep}B, we present some patterns
  representing alternative parameter choices. Fig.\,\ref{fig:paramsweep}B i)
  shows the simulation result with low diffusion constant, $D$. Reducing the
  diffusion constant is equivalent to increasing the size of the domain $S$
  and so this simulation can be thought to represent an animal with a large
  cortical area but with unchanged cortical complexity. In this simulation, a
  large proportion of the domain is occupied by the projections with
  intermediate interaction parameters; those with strong interaction
  parameters are bunched up towards the edge of the domain. A barrel pattern
  fails to form. Fig.\,\ref{fig:paramsweep}B ii)~and iii)~indicate the effect
  of varying $D$ in either direction from its optimum. ii)~is a less
  exaggerated version of the pattern in i), whereas iii)~shows an expansion of
  the outer barrels and a corresponding compression of the barrels within.}

\cmnt{We performed a sensitivity analysis to determine if the quality of the
  pattern was sensitive to any of the following inputs: i)~The angle and
  magnitude of each of the two guidance fields (Fig.\,\ref{fig:main}A);
  ii)~the magnitude, offset and characteristic length scale of the noise
  applied to $a_i$ at $t=0$; and iii)~the magnitude of each of the interaction
  parameters, $\gamma_{i,j}$. We analysed the effect on the three pattern
  metrics, $\delta$, $\eta$ and $\omega$ evaluated after 30000 simulation
  steps. First, we conducted 10 runs of the simulation to evaluate the mean
  and standard deviation (as a result of the different random initial
  conditions for $a_i$) of the three metrics in the default model
  (Fig.\,\ref{fig:main}C--E) which were $\delta=0.089\pm{0.004}$,
  $\eta=0.531\pm{0.005}$ and $\omega=0.2165\pm{0.0001}$). These results are
  not strongly affected by the random initial conditions for $a_i$ but give a
  baseline from which to evaluate whether changes in the other inputs have a
  statistically significant effect.}

@@

The present results suggest that the key requirements for the emergence of
realistic barrel patterning are i)~at each cortical location thalamocortical
projections compete for a limited number of available synaptic connections
(Eqs.~\ref{eq:dc}--\ref{eq:da}), ii)~at each location the branching rate of a
given projection is reduced by the density of other projections
(Eq.\,\ref{eq:comp}), and iii)~the branch density of each projection is
conserved over time.
